Skill estimation using TrueSkill Through Time: \\ the \proglang{Julia}, \proglang{Python} and \proglang{R} packages (Draft)

### 2021-06-14 doc/article.tex
Submitted version

\subsection{\en{Sequence of events}\es{Secuencia de eventos}} \label{sec:sequence_of_events}

@@

\subsection{\en{The history of the Association of Tennis Professionals (ATP)}\es{Historia de la Asociación de Tenistas Profesionales (ATP)}} \label{sec:atp}

@@

% Parrafo
\en{ATP's ranking points are updated every Monday according to the prestige of the tournament and the stage reached. }%
\es{Los puntos de la clasificación de la ATP se actualizan semanalmente en función del prestigio del torneo y la fase alcanzada. }%
%
\en{Only during brief periods there is no coincidence between the estimated skill and the player who is at the top of the ATP's ranking, showing a good agreement between both methodologies. }%
\es{Hay una buena concordancia entre ambas metodologías respecto de la detección de quién es el mejor jugador en cada momento. }%
%
\en{The ATP's ranking take time to identify the best player because the points are dropped a year after being awarded. }%
\es{La clasificación de la ATP tarda en identificar al mejor jugador debido que su método retira los puntos un año después de ser otorgados. }%
\en{Perhaps more interesting are the differences with respect to the historical ranking shown in Table~\ref{tab:atp_weeks}. }%
\es{Quizás más interesante sean las diferencia respecto del ranking histórico que se muestra en la tabla~\ref{tab:atp_weeks}. }%
%\en{Notwithstanding, there are some differences between players' skills and the ATP's ranking, especially concerning the historical ranking shown in Table~\ref{tab:atp_weeks}. }%
%
\en{For example, Lleyton Hewitt's position in the historical ranking is the product of a window of opportunity opened around the year 2000, since his ability is relatively low in historical terms. }%
\es{La posición que obtiene Lleyton Hewitt en el ranking histórico de la ATP es producto de una ventana de oportunidad abierta cerca del año 2000, pues su habilidad es relativamente baja en términos históricos. }% 
%
\en{On the other hand, Andy Murray is the fourth most skilled player according to our estimations, but he only reaches 14th place in the historical ranking, just one place above Ilie N\u{a}stase. }%
\es{Por otro lado, Andy Murray, siendo el cuarto jugador más hábil, sólo ocupa el puesto 14 en la clasificación histórica de la ATP, sólo un lugar por encima de Ilie N\u{a}stase. }%

@@

h_ground = History(composition = composition_ground, times = dates, sigma = 1.0, gamma = 0.01, beta = 0.0, priors = priors)
convergence(h_ground, epsilon = 0.01, iterations=10)
\en{In this example we keep the same prior as before for all the generic players, but this time we define them using the variable \texttt{priors}. }%
\es{En este ejemplo mantenemos el mismo prior para todos los jugadores genéricos, pero en esta ocasión los definidos usando la variable \texttt{priors}. }%
\en{Just as in the previous code (\ref{lst:atp}), we create the teams depending on whether the game is double or single, but now adding the specific skills of each player as their teammate (we use the operator \texttt{*} to concatenate strings). }%
\es{Al igual que en el código anterior (\ref{lst:atp}), creamos los equipos en función de si la partida es doble o simple, pero ahora añadiendo las habilidades específicas de cada jugador como su compañero de equipo (usamos el operador \texttt{*} para concatenar strings). }%
\en{As the specific skills are not defined in the dictionary \texttt{prior}, they will be initialized with the default values indicated in the class \texttt{History}. }%
\es{Como las habilidades específicas no están definidas en el diccionario \texttt{prior}, ellas van a ser inicializadas con los valores por defecto que le indicamos a la clase \texttt{History}. }%
\en{For those factors we chose a null \texttt{beta} so as not to add more noise to the players' performance, keeping the scale of the estimates stable. }%
\es{Para ellos elegimos un \texttt{beta} nulo de modo tal de no agregar más ruido al rendimiento de los jugadores, manteniendo estable la escala de las estimaciones. }%
\en{We choose a \texttt{sigma} that we consider sufficiently large and a dynamic factor \texttt{gamma} that represents \SI{1}{\percent} of the a priori uncertainty. }%
\es{Elegimos un \texttt{sigma} que consideramos suficientemente grande y una factor dinámico \texttt{gamma} que represente el 1\% de la incertidumbre a priori. }%
    \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[page={1},width=\linewidth]{figures/atp_ground}
    \caption{Nadal}
    \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[page={3},width=\linewidth]{figures/atp_ground}
    \caption{Djokovic}
    \en{Skill difference on the three main types of ground. }%
    \es{Diferencia de habilidad en los tres tipos de suelo principales. }%
    \en{Each point on the $y$-axis represents a distance of one $\beta$, i.e. 76\% probability of winning. }%
    \es{Cada punto del eje $y$ representa una $\beta$ de distancia, es decir 76\% de probabilidad de ganar. }%
    \label{fig:atp_ground}
\en{In Figure~\ref{fig:atp_ground} we show the skill difference that Nadal and Djokovic have in each of the three types of ground. }%
\es{En la figura~\ref{fig:atp_ground} mostramos la diferencia de habilidad que Nadal y Djokovic tiene entre los tres tipos de piso. }%
\en{We can see that Nadal has a big skill difference between grounds, unlike Djokovic who has very similar skills on all three types of ground. }%
\es{Podemos ver que Nadal tiene una diferencia de habilidad grande entre suelos, a diferencia de Djokovic que tiene habilidades muy similares en los tres tipos de suelo. }%
\en{The Nadal's skill difference between clay and grass grounds is greater than one $\beta$, which means at least 76\% difference in probability of winning compared to itself. }%
\es{La diferencia de habilidad de Nadal entre piso de polvo de ladrillo y pasto es mayor a un $\beta$, lo que significa una diferencia de al menos 76\% de probabilidad de ganar si lo comparamos consigo mismo. }%
\en{In the case of Nadal it seems important to model the skill's multi-dimensionality, while in Djokovic's case it seems reasonable to summarize it in a single dimension. }%
\es{En el caso de Nadal parece importante modelar la multi-dimensionalidad de la habilidad, mientras que en el caso de Djokovic parece razonable resumir la habilidad en una única dimensión. }%
\en{To assess whether the complexity added by modeling multidimensionality is appropriate in general terms, we can compare the joint prior prediction of the models, calling the method \texttt{log\_evidence()} of the class \texttt{History}. }%
\es{Para evaluar si la complejidad que se agrega al modelar la multidimensional conviene en términos generales, podemos comparar la predicción a priori conjunta de los modelos, llamando al método \texttt{log\_evidence()} de la clase \texttt{History}. }%
\en{In tennis it is sufficient to summarize the skills in a single dimension, since the a prior prediction is maximized when the parameters of the ground factors, $\sigma$ and $\gamma$, vanish. }%
\es{En tenis es suficiente resumir las habilidades en una única dimensión, dado que la predicción a priori se maximiza cuando los parámetros de los factores de piso, $\sigma$ y $\gamma$, se desvanecen. }%
\en{In other examples, where the multidimensionality of skills is more relevant, it may be appropriate to separate the skills of all agents into different components. }%
\es{En otros ejemplos, donde la multidimensionalidad de las habilidades sea más relevante, convendrá separar la habildiad de todos los agente en diferentes componentes. }%
\en{If we consider only the games in which Nadal participates, optimality is achieved when the parameters take the values $\sigma=0.35$ and $\gamma=0$, meaning that it is necessary to model multidimensional skills ($\sigma>0$) but considering that their effect does not change over time ($\gamma = 0$). }%
\es{Si consideramos sólo las partidas en las que participa Nadal, la optimización se alcanza cuando los parámetros toman los valores $\sigma=0.35$ y $\gamma=0$, indicando que conivene modelar habilidades multidimensionales ($\sigma>0$) pero considerando que su efecto no cambia en el tiempo ($\gamma = 0$). }%
\en{In this scenario, Nadal's ability on Clay is $0.87\beta$ higher than on Hard, and $1.05\beta$ higher than on Grass. }%
\es{En este escenario, la habilidad de Nadal en polvo de ladrillo es $0.87\beta$ mayor a la de cemento, y $1.05\beta$ mayor a la de pasto. }%

@@

\en{In this section we provide the complete mathematical documentation of the TrueSkill Through Time model. }%
\es{En esta secci\'on ofrecemos la documentación matemática completa del modelo TrueSkill Through Time. }%
%
\en{The advantage of this model lies in the strict application of probability theory: all the assumptions are made explicit through a generative model, and the inference is solved only with the rules of probability, nothing more than the sum and the product rules. }%
\es{La ventaja de este modelo reside en la aplicación estricta de la teoría de la probabilidad: todos los supuestos se hacen expl\'icitos a trav\'es de un modelo generativo, y la inferencia se resuelve sólo con las regla de la probabilidad, nada más que las reglas de la suma y el producto. }%
%
\en{In section~\ref{sec:sumProductAlgorithm} we introduce the \emph{sum-product algorithm}, which allows us to efficiently apply this rules to compute the marginal distributions, e.g. the posterior and the prior prediction. }%
\es{En la secci\'on~\ref{sec:sumProductAlgorithm} introducimos el \emph{sum-product algorithm}, que nos permite aplicar eficientemente estas reglas para computar las distribuciones marginales, e.g. el posterior y la predicción a priori. }%
%
\en{In section~\ref{sec:propiedades} we list the properties that we will need to derive the marginal distributions of interest. }%
\es{En la secci\'on~\ref{sec:propiedades} enumeramos las propiedades que necesitaremos para derivar las distribuciones marginales de inter\'es. }%
%
\en{In section \ref{sec:Gaussian} we introduce the operations of the class \texttt{Gaussian}, which does most of the computation. }%
\es{En las sección \ref{sec:Gaussian} introducimos las operaciones de la clases \texttt{Gaussian}, la que realiza la mayor parte del c\'omputo. }%
%
\en{In sections \ref{sec:2vs2}, \ref{sec:empate}, and \ref{sec:approximate_posterior} we show how to solve the prior prediction and the exact posterior of an event, we introduce the darw model, we explain how to approximate the exact posterior in events with two teams. }%
\es{En las secciones \ref{sec:2vs2}, \ref{sec:empate}, \ref{sec:approximate_posterior}, mostramos c\'omo resolver la predicci\'on a priori y el posterior exacto de un evento, incorporamos empate al modelo y explicamos c\'omo aproximar el posterior exacto en partidas con dos equipos. }%
%
\en{In section \ref{sec:iterative_posterior} we explain the general multi-team solution, which requires the application of an iterative algorithm. }%
\es{En la sección \ref{sec:iterative_posterior} explicamos la soluci\'on general multi-equipos, que requiere la aplicación de un algoritmo iterativo. }%
%
\en{In section~\ref{sec:throughTime} we justify the mathematical steps required to solve the full TrueSkill Through Time model. }%
\es{En la secci\'on~\ref{sec:throughTime} justificamos los pasos matemáticos requeridos para resolver el modelo TrueSkill Through Time completo. }%

\subsection{Sum-product algorithm} \label{sec:sumProductAlgorithm}

\en{The \emph{sum-product algorithm}~\citep{Kschischang2001} takes advantage of the structure of the joint probability distribution, imposed by the causal model, to efficiently apply the rules of probability. }%
\es{El \emph{sum-product algorithm}~\citep{Kschischang2001} aprovecha la estructura de la distribuci\'on de probabilidad conjunta, impuesta por el modelo causal, para aplicar eficientemente las reglas de la probabilidad. }%
%
\en{Any model can be factored into the product of conditional probabilities. }%
\es{Cualquier modelo puede factorizarse en el producto de probabilidades condicionales. }%
%
\en{By making use of the independencies between variables, our model (figure~\ref{fig:generative_model}) can be factorized as,}
\es{Haciendo uso de las independencias entre las variables, nuestro modelo (figura~\ref{fig:generative_model}) puede factorizarse como,}
%
\begin{equation} \label{eq:factorization}
 p(\bm{s},\bm{p},d,r) = p(s_1)p(s_2)p(p_1|s_1)p(p_2|s_2)p(d|\bm{p})P(r|d)
\end{equation}
%
\en{In figure~\ref{fig:factor_graph} we show this factorization graphically. }%
\es{En la figura~\ref{fig:factor_graph} mostramos esta factorizaci\'on graficamente. }%
%
\en{These types of representations, known as \emph{factor graph}, are graphs with two types of nodes: variable nodes (white circles), and function nodes (black squares). }%
\es{Este tipo de representaciones, conocidas como \emph{factor graph}, son gr\'afos con dos tipos de nodos: nodos variables (círculos blancos), y nodos funciones (cuadrados negros). }%
%
\en{The edge between node variables and node functions represent the mathematical relationship ``the variable $v$ is an argument of the function $f$''. }%
\es{Los ejes entre nodos variables y nodos funciones representan la relaci\'on matem\'atica ``la variable $v$ es argumento de la funci\'on $f$''. }%
%
\begin{figure}[ht!]
\centering \small
    \tikz{         
%         \node[const, above=of fr] (nfr) {$f_r$}; %
% 	\node[const, above=of nfr] (dfr) {\large $\mathbb{I}(d >0)$}; %

@@

     \en{Graphical way of representing the factorization of joint distibution induced by the basic causal model (Eq.~\ref{eq:factorization}). }%

@@

     \en{Black squares represent the functions, white circles represent the variable, and the edges between them represent the mathematical relationship ``the variable is argument of the function''. }%

@@

\en{In our case we want to compute two marginals, the proportional posterior of the skills $p(s_i, r)$ and the a prior probability of the result $p(r)$. }%

@@

\en{The \emph{sum-product algorithm} is a general way of breaking down the rules of probability as messages that are sent locally between the nodes of the \emph{factor graph}. }%

@@

\en{There are two types of messages: the messages that variable nodes send to their functions neighbors ($m_{v \rightarrow f}(v)$); and the messages that function nodes send to their variable neighbors ($m_{f \rightarrow v}(v)$). }%
\es{Hay dos tipos de mensajes: los mensajes que envian los nodos variables a sus funciones vecinas ($m_{v \rightarrow f}(v)$); y los mensajes que envian los nodos funciones a sus variables vecinas ($m_{f \rightarrow v}(v)$). }%
\en{The former encodes a portion of the product rule. }%

@@

\en{Where $n(v)$ represents the set of node neighbors of $v$. }%
\es{Donde $n(v)$ representa el conjunto de vecinos del nodo $v$. }%
\en{In short, the messages sent by a $v$ variable is simply the product of the messages that $v$ received from the rest of their neighbors $h \in n(v)$ except $f$. }%

@@

\en{The messages sent by the function nodes encodes a portion of the sum rule. }%

@@

\en{Where $\bm{h} = n(f)\setminus \{v\}$ is the set of all of neighbors of $f$ except $v$, and $f(\bm{h},v)$ represents the function $f$, evaluated in all its arguments. }%
\es{Donde $\bm{h} = n(f)\setminus \{v\}$ es el conjunto de todos los vecinos de $f$ salvo $v$, y $f(\bm{h},v)$ represeta la funci\'on $f$, evaluada en todos sus argumentos. }%
\en{In short, the messages sent by a function $f$ to a neighboring variable $v$ is simply the integration (or sum) over $\bm{h}$ of the product of itself and all the messages that $f$ receives from the rest of its neighbors $\bm{h}$ except $v$. }%

@@

\subsection{\en{Mathematical properties and notation}\es{Propiedades matem\'aticas y notaci\'on}}\label{sec:propiedades}

\en{The efficiency of TrueSkill Through Time is achieved because the marginals can be computed analytically. }%
\es{La eficiencia de TrueSkill Through Time se obtiene gracias que las marginales se pueden computar de forma analítica. }%
%
\en{In this section we list the properties that we need to derive the exact and the approxiamte messages that arise from the sum-product algorithm. }%
\es{En esta secci\'on enumeramos las propiedades que necesitamos para derivar los mensajes exactos y aproximados que surgen del \emph{sum-product algorithm}. }%
%
\en{The first property states that the product of two Gaussian distributions, both evaluated at the same point $x$, can be expressed as the product of two other Gaussian distributions with only one of them evaluated at $x$. }%
\es{La primera propiedad establece que el producto de dos distribuciones gaussianas, ambas evaluadas en el mismo punto $x$, pueden expresarse como la producto de otras dos distribuciones gaussianas con sólo una de ellas evaluada en $x$. }%
%
\begin{equation*}\label{eq:gaussian_product} \tag{\text{\en{Gaussian product}\es{Producto de gaussianas}}}
\N(x|\mu_1,\sigma_1^2)\N(x|\mu_2,\sigma_2^2) \overset{\ref{multiplicacion_normales}}{=} \N(\mu_1|\mu_2,\sigma_1^2+\sigma_2^2) \N(x|\mu_{*},\sigma_{*}^2)
\end{equation*}
%
\en{where}\es{con} $\mu_{*} = \frac{\mu_1}{\sigma_1^2} + \frac{\mu_2}{\sigma_2^2}$ y $\sigma_{*}^2 = \left(\frac{1}{\sigma_1^2} + \frac{1}{\sigma_2^2} \right)^{-1}$.
%
\en{Something similar occurs with the division of two Gaussian distributions, both evaluated at the same point $x$. }%
\es{Algo similar ocurre con la división de dos distribuciones gaussianas, ambas evaluadas en el mismo punto $x$. }%
\begin{equation*}\label{eq:gaussian_division} \tag{\text{\en{Gaussian division}\es{División de gaussianas}}}
\N(x|\mu_1,\sigma_1^2)/\N(x|\mu_2,\sigma_2^2) \overset{\ref{sec:division_normales}}{\propto} \N(x|\mu_{\div},\sigma_{\div}^2)/\N(\mu_1|\mu_2,\sigma_1^2+\sigma_2^2) 
\end{equation*}
%
\en{where}\es{con} $\mu_{\div} = \frac{\mu_1}{\sigma_1^2} - \frac{\mu_2}{\sigma_2^2}$ y $\sigma_{\div}^2 = \left(\frac{1}{\sigma_1^2} - \frac{1}{\sigma_2^2} \right)^{-1}$.

>>>>

\subsection{\en{Sequence of Events}\es{Secuencia de eventos}} \label{sec:sequence_of_events}

@@

\subsection{\en{The History of the Association of Tennis Professionals (ATP)}\es{Historia de la Asociación de Tenistas Profesionales (ATP)}} \label{sec:atp}

@@

% Parrafo
\en{ATP's ranking points are updated every Monday according to the prestige of the tournament and the stage reached. }%
\es{Los puntos de la clasificación de la ATP se actualizan semanalmente en función del prestigio del torneo y la fase alcanzada. }%
%
\en{Only during brief periods there is no coincidence between the estimated skill and the player who is at the top of the ATP's ranking, showing a good agreement between both methodologies. }%
\es{Sólo durante cortos períodos de tiempo no hay concordancia entre las estimaciones de habilidad y quién está en el primer puesto del ranking ATP, mostrando una buen acuerdo entre ambas metodologías. }%
%
\en{Notwithstanding, there are some notorious differences between estimated players' skills and the ATP's ranking, especially concerning the historical ranking shown in Table~\ref{tab:atp_weeks}. }%
\es{Sin embargo, hay algunas diferencias notables con respecto al ranking histórico que se muestra en la tabla~\ref{tab:atp_weeks}. }%
%
\en{On the one hand, Lleyton Hewitt's position in the historical ranking is the product of a window of opportunity opened around the year 2000, since his ability is relatively low in historical terms. }%
\es{La posición que obtiene Lleyton Hewitt en el ranking histórico de la ATP es producto de una ventana de oportunidad abierta cerca del año 2000, pues su habilidad es relativamente baja en términos históricos. }% 
%
\en{On the other hand, Andy Murray is the fourth most skilled player according to our estimations, but he only reaches 14th place in the historical ranking, just one place above Ilie N\u{a}stase. }%
\es{Por otro lado, Andy Murray, siendo el cuarto jugador más hábil, sólo ocupa el puesto 14 en la clasificación histórica de la ATP, sólo un lugar por encima de Ilie N\u{a}stase. }%

@@

h_ground = History(composition = composition_ground, times = dates, sigma = 1.0, gamma = 0.01, beta = 0.0, priors = priors)
convergence(h_ground, epsilon = 0.01, iterations=10)
\end{lstlisting}
%
\en{In this example, we keep the same prior as in code~\ref{lst:atp} for all the generic players, but in this code, we define them using the variable \texttt{priors}. }%
\es{En este ejemplo mantenemos el mismo prior para todos los jugadores genéricos, pero en esta ocasión los definidos usando la variable \texttt{priors}. }%
%
\en{We create the teams depending on whether the game is double or single similarly to code~\ref{lst:atp}, but now adding the specific surface skills of each player as their teammate (we use the operator \texttt{*} to concatenate strings). }%
\es{Al igual que en el código anterior (\ref{lst:atp}), creamos los equipos en función de si la partida es doble o simple, pero ahora añadiendo las habilidades específicas de cada jugador como su compañero de equipo (usamos el operador \texttt{*} para concatenar strings). }%
%
\en{As the specific surface skills are not defined in the variable \texttt{prior}, they are initialized using the default values defined in the class \texttt{History}. }%
\es{Como las habilidades específicas no están definidas en el diccionario \texttt{prior}, ellas van a ser inicializadas con los valores por defecto que le indicamos a la clase \texttt{History}. }%
%
\en{We also define \texttt{beta} as null for specific surface skills to avoid adding additional noise to the players' performance, keeping the scale of the estimates stable. }%
\es{Para ellos elegimos un \texttt{beta} nulo de modo tal de no agregar más ruido al rendimiento de los jugadores, manteniendo estable la escala de las estimaciones. }%
%
\en{We select a \texttt{sigma} that we consider sufficiently large and a dynamic factor \texttt{gamma} representing \SI{1}{\percent} of the prior uncertainty. }%
\es{Elegimos un \texttt{sigma} que consideramos suficientemente grande y una factor dinámico \texttt{gamma} que represente el 1\% de la incertidumbre a priori. }%
\begin{figure}[ht!]
    \centering
    \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[page={1},width=\linewidth]{figures/atp_ground}
    \caption{Nadal}
    \end{subfigure}
    \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[page={3},width=\linewidth]{figures/atp_ground}
    \caption{Djokovic}
    \end{subfigure}
    \caption{
    \en{Skill difference on the three main types of surface. }%
    \es{Diferencia de habilidad en los tres tipos de suelo principales. }%
    %
    \en{Each point on the $y$-axis represents a distance of one $\beta$, i.e. 76\% probability of winning. }%
    \es{Cada punto del eje $y$ representa una $\beta$ de distancia, es decir 76\% de probabilidad de ganar. }%
    }
    \label{fig:atp_ground}
\end{figure}

%
\en{In Figure~\ref{fig:atp_ground}, we show the skill difference that Nadal and Djokovic have in each of the three types of ground. }%
\es{En la figura~\ref{fig:atp_ground} mostramos la diferencia de habilidad que Nadal y Djokovic tiene entre los tres tipos de piso. }%
%
\en{Nadal has a notorious skill difference when playing on different surfaces. On the contrary, Djokovic has very similar skills in the three types. }%
\es{Podemos ver que Nadal tiene una diferencia de habilidad grande entre suelos, a diferencia de Djokovic que tiene habilidades muy similares en los tres tipos de suelo. }%
%
\en{The Nadal's skill difference between clay and grass grounds is greater than one $\beta$, which means at least 76\% difference in probability of winning compared to itself. }%
\es{La diferencia de habilidad de Nadal entre piso de polvo de ladrillo y pasto es mayor a un $\beta$, lo que significa una diferencia de al menos 76\% de probabilidad de ganar si lo comparamos consigo mismo. }%
%
\en{In the case of Nadal, it seems important to model the skill's multi-dimensionality, while in Djokovic's case it seems reasonable to summarize it in a single dimension. }%
\es{En el caso de Nadal parece importante modelar la multi-dimensionalidad de la habilidad, mientras que en el caso de Djokovic parece razonable resumir la habilidad en una única dimensión. }%
%
\en{To assess whether the complexity added by modeling multi-dimensionality is appropriate in general terms, we can compare the joint prior prediction of the models, calling the method \texttt{log\_evidence()} of the class \texttt{History}. }%
\es{Para evaluar si la complejidad que se agrega al modelar la multidimensional conviene en términos generales, podemos comparar la predicción a priori conjunta de los modelos, llamando al método \texttt{log\_evidence()} de la clase \texttt{History}. }%

%Parrafo

\en{In tennis, it is sufficient to summarize the skills in a single dimension since the prior prediction is maximized when the parameters of the surface's factors (i.e. $\sigma$ and $\gamma$) vanish. }%
\es{En tenis es suficiente resumir las habilidades en una única dimensión, dado que la predicción a priori se maximiza cuando los parámetros de los factores de piso, $\sigma$ y $\gamma$, se desvanecen. }%
%
\en{In other examples, where the multi-dimensionality of skills could be more relevant, it should be necessary to model the skills of all agents using different components. }%
\es{En otros ejemplos, donde la multidimensionalidad de las habilidades sea más relevante, convendrá separar la habildiad de todos los agente en diferentes componentes. }%
%
\en{If we consider only the games in which Nadal participates, optimality is achieved when the parameters take the values $\sigma=0.35$ and $\gamma=0$, meaning that it is necessary to model multidimensional skills ($\sigma>0$) but considering that their effect does not change over time ($\gamma = 0$). }%
\es{Si consideramos sólo las partidas en las que participa Nadal, la optimización se alcanza cuando los parámetros toman los valores $\sigma=0.35$ y $\gamma=0$, indicando que conivene modelar habilidades multidimensionales ($\sigma>0$) pero considerando que su efecto no cambia en el tiempo ($\gamma = 0$). }%
%
\en{In this scenario, Nadal's ability on Clay is $0.87\beta$ higher than on Hard and $1.05\beta$ higher than on Grass. }%
\es{En este escenario, la habilidad de Nadal en polvo de ladrillo es $0.87\beta$ mayor a la de cemento, y $1.05\beta$ mayor a la de pasto. }%

@@

\en{In this section, we provide the complete mathematical documentation of the TrueSkill Through Time model. }%
\es{En esta secci\'on ofrecemos la documentación matemática completa del modelo TrueSkill Through Time. }%
%
\en{The advantage of this model lies in the strict application of probability theory: all the assumptions are made explicit through a generative model, and the inference is solved only with the rules of probability, nothing more than the sum and the product rules. }%
\es{La ventaja de este modelo reside en la aplicación estricta de la teoría de la probabilidad: todos los supuestos se hacen expl\'icitos a trav\'es de un modelo generativo, y la inferencia se resuelve sólo con las regla de la probabilidad, nada más que las reglas de la suma y el producto. }%
%
\en{In section~\ref{sec:sumProductAlgorithm}, we introduce the \emph{sum-product algorithm}, which allows us to apply these rules to compute the marginal distributions, e.g. the posterior and the prior prediction. }%
\es{En la secci\'on~\ref{sec:sumProductAlgorithm} introducimos el \emph{sum-product algorithm}, que nos permite aplicar eficientemente estas reglas para computar las distribuciones marginales, e.g. el posterior y la predicción a priori. }%
%
\en{In section~\ref{sec:propiedades}, we list the properties that we will need to derive the marginal distributions of interest. }%
\es{En la secci\'on~\ref{sec:propiedades} enumeramos las propiedades que necesitaremos para derivar las distribuciones marginales de inter\'es. }%
%
\en{In section \ref{sec:Gaussian}, we introduce the operations of the class \texttt{Gaussian}, which does most of the computation. }%
\es{En las sección \ref{sec:Gaussian} introducimos las operaciones de la clases \texttt{Gaussian}, la que realiza la mayor parte del c\'omputo. }%
%
\en{In sections \ref{sec:2vs2}, \ref{sec:empate}, and \ref{sec:approximate_posterior}, we show how to solve the prior prediction and the exact posterior of an event, we introduce the model for a draw, and we explain how to approximate the exact posterior in events with two teams. }%
\es{En las secciones \ref{sec:2vs2}, \ref{sec:empate}, \ref{sec:approximate_posterior}, mostramos c\'omo resolver la predicci\'on a priori y el posterior exacto de un evento, incorporamos empate al modelo y explicamos c\'omo aproximar el posterior exacto en partidas con dos equipos. }%
%
\en{In section \ref{sec:iterative_posterior}, we explain the general multi-team solution which requires using an iterative algorithm. }%
\es{En la sección \ref{sec:iterative_posterior} explicamos la soluci\'on general multi-equipos, que requiere la aplicación de un algoritmo iterativo. }%
%
\en{In section~\ref{sec:throughTime}, we justify the mathematical steps required to solve the full TrueSkill Through Time model. }%
\es{En la secci\'on~\ref{sec:throughTime} justificamos los pasos matemáticos requeridos para resolver el modelo TrueSkill Through Time completo. }%

\subsection{Sum-product Algorithm} \label{sec:sumProductAlgorithm}

\en{The \emph{sum-product algorithm}~\citep{Kschischang2001} takes advantage of the structure of the joint probability distribution, imposed by the causal model, to efficiently apply the rules of probability. }%
\es{El \emph{sum-product algorithm}~\citep{Kschischang2001} aprovecha la estructura de la distribuci\'on de probabilidad conjunta, impuesta por el modelo causal, para aplicar eficientemente las reglas de la probabilidad. }%
%
\en{Any model can be factorized into the product of conditional probabilities. }%
\es{Cualquier modelo puede factorizarse en el producto de probabilidades condicionales. }%
%
\en{As the variables are independent, our model (figure~\ref{fig:generative_model}) can be factorized as,}
\es{Haciendo uso de las independencias entre las variables, nuestro modelo (figura~\ref{fig:generative_model}) puede factorizarse como,}
%
\begin{equation} \label{eq:factorization}
 p(\bm{s},\bm{p},d,r) = p(s_1)p(s_2)p(p_1|s_1)p(p_2|s_2)p(d|\bm{p})P(r|d)
\end{equation}
%
\en{In Fig.~\ref{fig:factor_graph}, we show this factorization graphically. }%
\es{En la figura~\ref{fig:factor_graph} mostramos esta factorizaci\'on graficamente. }%
%
\en{These representations, known as \emph{factor graphs}, have two kinds of nodes: variable-type nodes (white circles) and function-type nodes (black squares). }%
\es{Este tipo de representaciones, conocidas como \emph{factor graph}, son gr\'afos con dos tipos de nodos: nodos variables (círculos blancos), y nodos funciones (cuadrados negros). }%
%
\en{The edge between node variables and node functions represents the mathematical relationship ``the variable $v$ is an argument of the function $f$''. }%
\es{Los ejes entre nodos variables y nodos funciones representan la relaci\'on matem\'atica ``la variable $v$ es argumento de la funci\'on $f$''. }%
%
\begin{figure}[ht!]
\centering \small
    \tikz{         
%         \node[const, above=of fr] (nfr) {$f_r$}; %
% 	\node[const, above=of nfr] (dfr) {\large $\mathbb{I}(d >0)$}; %

@@

     \caption{%
     \en{Graphical way of representing the factorization of joint distribution induced by the basic causal model, represented by Eq.~\eqref{eq:factorization}. }%

@@

     \en{Black squares represent the functions, white circles represent the variables, and the edges between them represent the mathematical relationship ``the variable is the argument of the function''. }%

@@

\en{In our case, we want to compute two marginals, the proportional posterior of the skills $p(s_i, r)$ and the prior probability of the result $p(r)$. }%

@@

\en{The \emph{sum-product algorithm} is a general way of breaking down the rules of probability as messages locally sent between the nodes of the \emph{factor graph}. }%

@@

\en{There are two types of messages: those sent by variable-type nodes to their function-type neighbors ($m_{v \rightarrow f}(v)$) and the ones that function-type nodes send to their variable-type neighbors ($m_{f \rightarrow v}(v)$). }%
\es{Hay dos tipos de mensajes: los mensajes que envian los nodos variables a sus funciones vecinas ($m_{v \rightarrow f}(v)$); y los mensajes que env\'ian los nodos funciones a sus variables vecinas ($m_{f \rightarrow v}(v)$). }%
\en{The former partially performs the product rule. }%

@@

\en{where $n(v)$ represents the set of neighbor nodes to $v$. }%
\es{donde $n(v)$ representa el conjunto de vecinos del nodo $v$. }%
\en{In a brief, the messages sent by the variable-type node $v$ are simply the product of the messages that $v$ receives from the rest of their neighbors $h \in n(v)$ except $f$. }%

@@

\en{The messages sent by the function-type nodes encode a portion of the sum rule. }%

@@

\en{where $\bm{h} = n(f)\setminus \{v\}$ is the set of all neighbors to $f$ except $v$, and $f(\bm{h},v)$ represents the function $f$, evaluated in all its arguments. }%
\es{donde $\bm{h} = n(f)\setminus \{v\}$ es el conjunto de todos los vecinos de $f$ salvo $v$, y $f(\bm{h},v)$ representa la funci\'on $f$, evaluada en todos sus argumentos. }%
\en{The messages sent by the function-type node $f$ to a neighboring variable-type node $v$ is simply the integration (or sum) over $\bm{h}$ of the product of itself and all the messages that $f$ receives from the rest of its neighbors $\bm{h}$ except $v$. }%

@@

\subsection{\en{Mathematical Properties and Notation}\es{Propiedades matem\'aticas y notaci\'on}}\label{sec:propiedades}

\en{The efficiency of TrueSkill Through Time is based on the analytical computation of marginal probabilities. }%
\es{La eficiencia de TrueSkill Through Time se obtiene gracias que las marginales se pueden computar de forma analítica. }%
%
\en{In this section, we list the properties that we need to derive the exact and the approximate messages that arise from the sum-product algorithm. }%
\es{En esta secci\'on enumeramos las propiedades que necesitamos para derivar los mensajes exactos y aproximados que surgen del \emph{sum-product algorithm}. }%
%
\en{The first property states that the product of two Gaussian distributions, both evaluated at the same point $x$, can be expressed as the product of two other Gaussian distributions with only one of them evaluated at $x$. }%
\es{La primera propiedad establece que el producto de dos distribuciones gaussianas, ambas evaluadas en el mismo punto $x$, pueden expresarse como la producto de otras dos distribuciones gaussianas con sólo una de ellas evaluada en $x$. }%
%
\begin{equation*}\label{eq:gaussian_product} \tag{\text{\en{Gaussian product}\es{Producto de gaussianas}}}
\N(x|\mu_1,\sigma_1^2)\N(x|\mu_2,\sigma_2^2) \overset{\ref{multiplicacion_normales}}{=} \N(\mu_1|\mu_2,\sigma_1^2+\sigma_2^2) \N(x|\mu_{*},\sigma_{*}^2)
\end{equation*}
%
\en{where}\es{con} $\mu_{*} = \frac{\mu_1}{\sigma_1^2} + \frac{\mu_2}{\sigma_2^2}$ y $\sigma_{*}^2 = \left(\frac{1}{\sigma_1^2} + \frac{1}{\sigma_2^2} \right)^{-1}$.
%
\en{Something similar occurs with the division of two Gaussian distributions, both evaluated at the same point $x$. }%
\es{Algo similar ocurre con la división de dos distribuciones gaussianas, ambas evaluadas en el mismo punto $x$. }%
\begin{equation*}\label{eq:gaussian_division} \tag{\text{\en{Gaussian division}\es{División de gaussianas}}}
\N(x|\mu_1,\sigma_1^2)/\N(x|\mu_2,\sigma_2^2) \overset{\ref{sec:division_normales}}{\propto} \N(x|\mu_{\div},\sigma_{\div}^2)/\N(\mu_1|\mu_2,\sigma_1^2+\sigma_2^2) 
\end{equation*}
%
\en{where}\es{con} $\mu_{\div} = \frac{\mu_1}{\sigma_1^2} - \frac{\mu_2}{\sigma_2^2}$ y $\sigma_{\div}^2 = \left(\frac{1}{\sigma_1^2} - \frac{1}{\sigma_2^2} \right)^{-1}$.
